Smart Model Elimination for Efficient Automated Machine Learning in Healthcare

### 2024-06-24 paper/paper.tex
Update paper.tex

\begin{abstract}
  Automated Machine Learning or AutoML has emerged as a popular field of research. We performed a literature review of existing AutoML papers and conducted an in-depth survey on five popluar AutoML frameworks. Many of these frameworks optimize a form of model selection, however, they all require every model to be run and evaluated. We propose a novel framework, SMEML (Smart Model Elimination Machine Learning), that automatically eliminates models that are unlikely to be performant. SMEML demonstrates the ability to generate a model with near equal accuracy to a "dumb" brute-force model, but in a fraction of the time. We also believe this innovation is particularly relevant to machine learning in healthcare because of the necessity for efficient diesease prediction. 
\end{abstract}

@@

Machine learning (ML) is a type of artificial intelligence (AI) that uses searching and trend-determining to make predictions. ML algorithms are trained and tested on a given dataset to become "smarter". Once they've learned the patterns typical of that data, they can make predictions based on similar but unseen data. These algorithms also require additional processes, however, like hyperparamater tuning, and can be modified in conjunction with one another, like in ensemble learning. This makes the process complex and time-inefficient. To do this on a larger scale and at greater speeds, recent research and effort have been put into Auto ML frameworks [1].

@@

Keeping the factors we have laid out in mind, namely expeditious running, accuracy, and interpretability, we also uncovered a key continuity in framework design: frameworks are forced to run and train the majority of their available models to maximize accuracy at the expense of efficiency. Given the restricted nature of tabular data, there are some models that are less likely to perform well, thus running and training these models is frequently wasteful. Cutting these models completely, however, is not optimal either, as they are the best option in a minority of cases. We hypothesized that using a combination of layered machine learning implementation and meta-feature extraction would yield an equally versatile system that minimizes time spent on prediction while still maximizing the other quantities.

\section{Methodology}
When considering a specific dataset, various aspects regarding the distribution and relationship within the data can give us clues as to which models are likely to perform best. For example, linear models tend to work better on datasets where there is a linear relationship between features [11]. As a result, we can develop a boosting model to predict the best models based off extracted attributes from a dataset. These extracted attributes are shown in Table 2. Each attribute attempts to capture a specific type of relationship or distribution within the data. The variable importance of each attribute is shown in the "Weight" column. Linearity and data distribution are the most important attributes in determining the best model.

@@

To implement this model elimination process, we trained a boosting model on 268 kaggle datasets. We chose datasets that were medically related and focused our approach on binary classification. We extracted the attributes in Table 2 from each dataset: this is the input for the boosting model. We then ran a "dumb" framework that trained every model on each dataset. The accuracy of every model is ranked and every model is treated as a label with its value being its rank. Then we trained a multilabel regression to predict the rank of each model, giving the extracted attributes. We produced a median spearman correlation of 0.7301 between the predicted rank and the actual rank and a median p-value of 1.032e-05. Furthermore, we guarantee that we can predict a top two model within the predicted top eight 90.12\% of the time.

@@

Several observations were readily apparent from running these datasets. First, accuracy did not change for 4 of the 6 benchmarks. The other two benchmarks on eye states and diabetes prediction represented a loss of accuracy for SMEML by 0.13\% and 1.95\%, respectively. Training time, however, favored SMEML in every trial. SMEML outperformed the dummy framework by 173.96 seconds in its worst performance on the lung cancer trial and by 383.94 seconds in its best performance on the eye state dataset. Relative to the dummy's performance, a percent decrease in time to run for SMEML of at least 46.78\% could be seen, with a decrease of up to 69.22\% observerd as well. SMEML's worst performance by percent decrease was also the performance in which it returned perfect accuracy. 
\section{Discussion}
\subsection{Summary}
SMEML manages to return comparable accuracy results compared to its dumb couterpart while more than splitting the run time in half in the majority of benchmarks. Its performance is not limited by the size of the dataset, and it in fact returned its first, third, and fourth best percentage-based run times on the 'large' dataset benchmarks (those exceeding 1000 rows). This is important in third world medicine where time is of the essence and there is a constant influx of new patient data. Further research should be done in incorporating this innovation into a full-fledged AutoML framework to expedite and improve the process further. A practical demonsration of this software's real-world application can be found on our SmartDx website [18]. Source code can be found at [19].
\subsection{Limitations}
Our implementation of SMEML has limitations. These include a small training set due to lack of available datasets, finite computing power, time constraints, and a lack of access to paticular scientific literature. Our framework only improves on the model selection process and does not address the various other aspects of AutoML. Therefore, SMEML is not comparable to other full-featured AutoML frameworks.
\subsection{Future Work}
Future work will include increasing the size of the training set, implementing a fully fledged AutoML framework, and conducting a detailed experiment comparing SMEML to other popular AutoML frameworks. Furthermore, we can examine other types of machine learning such as regression and multiclass-classification.
\begin{ack}
The authors would like to thank the St. Mark's School of Texas for supporting this research.
\end{ack}

>>>>

\begin{abstract}
  Automated Machine Learning or AutoML has emerged as a popular field of research. We performed a literature review of existing AutoML papers and conducted a survey on five popular AutoML frameworks. Many of these frameworks optimize a form of model selection, but they all require every model to be run and evaluated. We propose a novel framework, SMEML (Smart Model Elimination Machine Learning), that automatically eliminates models that are unlikely to be performant. SMEML demonstrates the ability to generate a model with near equal accuracy to a "dumb" brute-force model, but in a fraction of the time. We also believe this innovation is particularly relevant to machine learning in healthcare because of the necessity for efficient diesease prediction. 
\end{abstract}

@@

Machine learning (ML) is a type of artificial intelligence (AI) that uses searching and trend-determining to make predictions. ML algorithms are trained and tested on a given dataset to become "smarter". Once they've learned the patterns typical of that data, they can make predictions based on similar data. These algorithms also require additional processes, however, like hyperparamater tuning, and can be modified in conjunction with one another, like in ensemble learning. This makes the process complex and time-inefficient. To do this on a larger scale and at greater speeds, recent research and effort have been put into Auto ML frameworks [1].

@@

Keeping the factors we have laid out in mind, we also uncovered a key continuity in framework design: frameworks must run the majority of their models to maximize accuracy at the expense of efficiency. Given the restricted nature of tabular data, there are some models that are less likely to perform well, thus running and training these models is wasteful. Cutting these models completely, however, is not optimal, as they are the best option in a minority of cases. We hypothesized that using a combination of layered machine learning implementation and meta-feature extraction would yield an equally versatile system that minimizes time spent on prediction while still maximizing the other quantities.

\section{Methodology}
When considering a specific dataset, the distribution and relationship between datapoints can give us clues as to which models are likely to perform best. For example, linear models tend to work better on datasets where there is a linear relationship between features [11]. Accordingly, we can develop a boosting model to predict the best models based off extracted attributes from a dataset. These extracted attributes are shown in Table 2. Each attribute attempts to capture a specific type of relationship or distribution within the data. The variable importance of each attribute is shown in the "Weight" column. Linearity and data distribution are the most important attributes in determining the best model.

@@

To implement this model elimination process, we trained a boosting model on 268 kaggle datasets. We chose medical datasets and focused our approach on binary classification. We extracted the attributes in Table 2 from each dataset: this is the input for the boosting model. We then ran a "dumb" framework that trained every model on each dataset. The accuracy of every model is ranked and every model is treated as a label with its value being its rank. Then we trained a multilabel regression to predict the rank of each model, giving the extracted attributes. We produced a median spearman correlation of 0.7301 between the predicted rank and the actual rank and a median p-value of 1.032e-05. Furthermore, we guarantee that we can predict a top two model within the predicted top eight 90.12\% of the time.

@@

Several observations were made from running these datasets. First, accuracy did not change for 4 of the 6 benchmarks. The other two benchmarks on eye states and diabetes prediction represented a loss of accuracy for SMEML by 0.13\% and 1.95\%, respectively. Training time, however, favored SMEML in every trial. SMEML outperformed the dummy framework by 173.96 seconds in its worst performance on the lung cancer trial and by 383.94 seconds in its best performance on the eye state dataset. Relative to the dummy's performance, a percent decrease in time to run for SMEML of at least 46.78\% could be seen, with a decrease of up to 69.22\% observerd as well. SMEML's worst performance by percent decrease was also the performance in which it returned perfect accuracy. 
\section{Discussion}
\subsection{Summary}
SMEML manages to return comparable accuracy results to its dumb couterpart while more than splitting the run time in half in the majority of benchmarks. Its performance is not limited by the size of the dataset, and it returned its first, third, and fourth best percentage-based run times on the 'large' dataset benchmarks (those exceeding 1000 rows). This is important in third world medicine where time is of the essence and there is a constant influx of new patient data. A practical demonsration of this software's real-world application can be found on our SmartDx website [18]. Source code can be found at [19].
\subsection{Limitations}
Our implementation of SMEML has limitations. These include a small training set due to lack of available datasets, finite computing power, time constraints, and a lack of access to paticular scientific literature. Our framework only improves on the model selection process and does not address the various other aspects of AutoML. Therefore, SMEML is not comparable to other full-featured AutoML frameworks.
\subsection{Future Work}
Future work will include increasing the size of the training set, implementing a fully fledged AutoML framework, and conducting a detailed experiment comparing SMEML to other popular AutoML frameworks. Furthermore, we can examine other types of machine learning such as regression and multiclass-classification.
\begin{ack}
The authors would like to thank the St. Mark's School of Texas for supporting this research.
\end{ack}
